---
title: "Additive Smoothing"
author: "Maurício Collaça"
date: "May 13, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

[https://en.wikipedia.org/wiki/Additive_smoothing]()

In statistics, additive smoothing, also called Laplace smoothing (not to be confused with Laplacian smoothing), or Lidstone smoothing, is a technique used to smooth categorical data. Given an observation $x = (x_1, ..., x_d)$ from a multinomial distribution with $N$ trials, a "smoothed" version of the data gives the estimator: $$\hat\theta_i = \frac{x_i + \alpha}{N + \alpha d}$$

Where the pseudocount $\alpha$ > 0 is the smoothing parameter.

```{r}
smoothing <- function(x, N, alpha) (x + alpha) / (N + alpha * length(x))
```

```{r}
set.seed(2)
x <- runif(10)
x
```
add-one smoothing with 1 trial
```{r}
y <- smoothing(x, 1, 1)
y
```
```{r}
plot.smoothing <- function(x, N, alpha) {
    y <- smoothing(x, N, alpha)
    df <- data.frame(x=rep(seq(x),2),
                     y=c(x,y),
                     series=rep(c("x",paste0("smoothing")), each=length(x)))
    ggplot(df, aes(x,y, color=series)) + geom_line() + geom_point()
}
plot.smoothing(x, N = 1, alpha = 1)
```

```{r}
set.seed(2)
x <- runif(10)
N <- 1:3
alpha <- c(0.1, 1, 10)
df <- expand.grid(x=x, N = N, alpha = alpha, KEEP.OUT.ATTRS = FALSE)
df$series <- rep(seq(length(N)*length(alpha)), each=length(x))
df$y <- unlist(lapply(split(df, df$series), function(sdf) smoothing(sdf$x, sdf$N, sdf$alpha)), use.names = F)
df$i <- unlist(lapply(split(df, df$series), function(sdf) seq_along(sdf$x)))
ggplot(df, aes(i,x)) +
    facet_grid(N~alpha, labeller="label_both", switch = "y", scales = "free") +
    theme(legend.position = "none") +
    geom_line() + geom_point() + geom_line(aes(i,y), color="red") + geom_point(aes(i,y), color="red")
```

$\alpha$ = 0 corresponds to no smoothing
```{r}
plot.smoothing(x, 1, 0)
```

