---
title: "Prediction testing"
author: "Maurício Collaça"
date: "April 24, 2018"
output: 
  html_document: 
    code_folding: hide
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "")
options(width = 120)
```

```{r message=FALSE}
library(data.table); library(quanteda)
```

# Predicting quiz 2

```{r}
quiz2 <- c("The guy in front of me just bought a pound of bacon, a bouquet, and a case of",
           "You're the reason why I smile everyday. Can you follow me please? It would mean the",
           "Hey sunshine, can you follow me and make me the",
           "Very early observations on the Bills game: Offense still struggling but the",
           "Go on a romantic date at the",
           "Well I'm pretty sure my granny has some old bagpipes in her garage I'll dust them off and be on my",
           "Ohhhhh #PointBreak is on tomorrow. Love that film and haven't seen it in quite some",
           "After the ice bucket challenge Louis will push his long wet hair out of his eyes with his little",
           "Be grateful for the good times and keep the faith during the",
           "If this isn't the cutest thing you've ever seen, then you must be")
```

## 1: The guy in front of me just bought a pound of bacon, a bouquet, and a case of *beer*
```{r}
rank.answers(quiz2[1], c("cheese", "soda", "beer", "pretzels"))
```
_Found @ 5gram if ordered by -(frequency*prob) or -frequency_
```{r}
predict.nextword(quiz2[1])
```

## 2: You're the reason why I smile everyday. Can you follow me please? It would mean the *world*
```{r}
rank.answers(quiz2[2], c("universe", "best", "most", "world"))
```
_Finds @ 5gram ordered by -prob, -frequency*prob or -frequency_
```{r}
predict.nextword(quiz2[2])
```

## 3: Hey sunshine, can you follow me and make me the *happiest*
```{r}
rank.answers(quiz2[3], c("saddest", "smellest", "happiest", "bluest"))
```
_Finds @ 5gram if ordered by -frequency*prob or -frequency_
```{r}
predict.nextword(quiz2[3])
```

## 4: Very early observations on the Bills game: Offense still struggling but the *defense*
```{r}
rank.answers(quiz2[4], c("players", "defense", "referees", "crowd"))
```
_Finds @ 4gram in the beginning of the sentence but cannot order to match the first_
```{r}
predict.nextword(quiz2[4])
```

## 5: Go on a romantic date at the *beach*
```{r}
rank.answers(quiz2[5], c("grocery", "movies", "mall", "beach"))
```
_Finds @ 4gram but cannot order to match the first_
```{r}
predict.nextword(quiz2[5])
```

## 6: Well I'm pretty sure my granny has some old bagpipes in her garage I'll dust them off and be on my *way*
```{r}
rank.answers(quiz2[6], c("phone", "horse", "motorcycle", "way"))
```
_Finds @ 5gram but cannot order to match the first_
_Finds @ 4gram if ordered by -frequency and not by -(frequency*order)_
```{r}
predict.nextword(quiz2[6])
```

## 7: Ohhhhh #PointBreak is on tomorrow. Love that film and haven't seen it in quite some *time*
```{r}
rank.answers(quiz2[7], c("time", "thing", "weeks", "years"))
```
_Finds @ 5gram as a unique response no matter the order_
```{r}
predict.nextword(quiz2[7])
```

## 8: After the ice bucket challenge Louis will push his long wet hair out of his eyes with his little *fingers*
```{r}
rank.answers(quiz2[8], c("toes", "eyes", "fingers", "ears"))
```
_Finds other answers @ 5gram beginning of sentence and 4 gram_
_Finds @ 3gram but cannot order to match the first_
```{r}
predict.nextword(quiz2[8])
```

## 9: Be grateful for the good times and keep the faith during the *bad*
```{r}
rank.answers(quiz2[9], c("sad", "hard", "bad", "worse"))
```
_Finds other answer at @ 4gram_
_Finds @ 4gram beginning of the sentence but cannot order to match the first_
```{r}
predict.nextword(quiz2[9])
```

## 10: If this isn't the cutest thing you've ever seen, then you must be *insane*
```{r}
rank.answers(quiz2[10], c("insane", "asleep", "callous", "insensitive"))
```
_Finds other answer @ 5gram and 4 gram_
_Finds @ 4gram beginning of the sentence but cannot order to match the first_
```{r}
predict.nextword(quiz2[10])
```

# Prediciting quiz 3

```{r}
quiz3 <- c("When you breathe, I want to be the air for you. I'll be there for you, I'd live and I'd",
           "Guy at my table's wife got up to go to the bathroom and I asked about dessert and he started telling me about his",
           "I'd give anything to see arctic monkeys this",
           "Talking to your mom has the same effect as a hug and helps reduce your",
           "When you were in Holland you were like 1 inch away from me but you hadn't time to take a",
           "I'd just like all of these questions answered, a presentation of evidence, and a jury to settle the",
           "I can't deal with unsymetrical things. I can't even hold an uneven number of bags of groceries in each",
           "Every inch of you is perfect from the bottom to the",
           "I’m thankful my childhood was filled with imagination and bruises from playing",
           "I like how the same people are in almost all of Adam Sandler's")
```


## 1: When you breathe, I want to be the air for you. I'll be there for you, I'd live and I'd *die*
```{r}
rank.answers(quiz3[1], c("give", "die", "sleep", "eat"))
```
```{r}
predict.nextword(quiz3[1])
```

## 2) Guy at my table's wife got up to go to the bathroom and I asked about dessert and he started telling me about his *spiritual*
```{r}
rank.answers(quiz3[2], c("horticultural", "financial", "marital", "spiritual"))
```
```{r}
predict.nextword(quiz3[2])
```

## 3) I'd give anything to see arctic monkeys this *weekend*
```{r}
rank.answers(quiz3[3], c("month", "decade", "morning", "weekend"))
```
```{r}
predict.nextword(quiz3[3])
```

## 4) Talking to your mom has the same effect as a hug and helps reduce your *stress*
```{r}
rank.answers(quiz3[4], c("hunger", "sleepiness", "happiness", "stress"))
```
```{r}
predict.nextword(quiz3[4])
```

## 5) When you were in Holland you were like 1 inch away from me but you hadn't time to take a *picture*
```{r}
rank.answers(quiz3[5], c("look", "minute", "picture", "walk"))
```
```{r}
predict.nextword(quiz3[5])
```

## 6) I'd just like all of these questions answered, a presentation of evidence, and a jury to settle the *matter*
```{r}
rank.answers(quiz3[6], c("matter", "case", "incident", "account"))
```
```{r}
predict.nextword(quiz3[6])
```

## 7) I can't deal with unsymetrical things. I can't even hold an uneven number of bags of groceries in each *hand*
```{r}
rank.answers(quiz3[7], c("finger", "hand", "toe", "arm"))
```
```{r}
predict.nextword(quiz3[7])
```

## 8) Every inch of you is perfect from the bottom to the *top*
```{r}
rank.answers(quiz3[8], c("side", "center", "top", "middle"))
```
```{r}
predict.nextword(quiz3[8])
```

## 9) I’m thankful my childhood was filled with imagination and bruises from playing *outside*
```{r}
rank.answers(quiz3[9], c("inside", "weekly", "daily", "outside"))
```
```{r}
predict.nextword(quiz3[9])
```

## 10) I like how the same people are in almost all of Adam Sandler's *movies*
```{r}
rank.answers(quiz3[10], c("stories", "pictures", "movies", "novels"))
```
```{r}
predict.nextword(quiz3[10])
```


# Predicting n words from a first one
```{r}
predict.nwords <- function(x, n) {
    for(tries in 1:n) {
        x <- paste(x, predict.nextword(x))
    }
    x   
}
predict.nwords("I",20)
```

# Predicting n random words
```{r}
predict.randomWords <- function(n) {
    predict.nwords(dt1[sample(1:nrow(dt1),1), x1],n-1)
}
predict.randomWords(20)
```

```{r}
predict.nextword("and")
```

